On the Programmatic Generation of Reproducible Documents

### 2020-07-15 inst/jss/listdown-jss.tex
Addressing JSS suggestions.

\author{
Michael Kane\\Yale University \And Simon Urbanek\\The University of
Aukland
}
\title{On the Programmatic Generation of Reproducible Documents}

\Plainauthor{Michael Kane, Simon Urbanek}
\Plaintitle{On the Programmatic Generation of Reproducible Documents}
\Shorttitle{On the Programmatic Generation of Reproducible Documents}

\Abstract{
Reproducible document standards, like \proglang{R} Markdown, facilitate
the programmatic creation of documents whose content is itself
programmatically generated. While these documents are generally not
complete in the sense that they will not include prose content,
generated by an author to provide context, a narrative, etc.,
programmatic generation can provide substantial efficiencies for
structuring and constructing documents. This paper explores the
programmatic generation of reproducible by distinguishing components
than can be created by computational means from those requiring
human-generated prose, providing guidelines for the generation of these
documents, and identifying a use case in clinical trial reporting. These
concepts and use case are illustrated through the \pkg{listdown} package
for the \proglang{R} programming environment, which is is currently
available on the Comprehensive R Archive Network (CRAN).
}

@@

@@

The \pkg{listdown} package provides functions to programmatically create
\proglang{R} Markdown files from named lists. It is intended for data
analysis pipelines where the presentation of the results is separated
from their creation. For this use case, a data processing (or analysis)
is performed and the results are provided in a single named list,
organized heirarchically. List element names denote sections,
subsections, subsubsection, etc. and the list elements contain the data
structure to be presented including graphs and tables. The package has
native support for \pkg{workflowr} \citep{blischak2019}, pdf, word, or
html document along with functions allowing a user to easily extend to
other types of supported documents. The goal of the package is to create
a documents with all tables and visualization that will appear
(computational components). This serves as a starting point from which a
user can organize outputs, describe a study, discuss results, and
provide conclusions (narrative components).

@@

One area where we have found \pkg{listdown} is particuarly useful is in
the reporting and research of clinical trial data. These collaborations
tend to be between (bio)statisticians and clinicians either analyzing
past trial data to formulate a new trial or in trial monitoring where
trial telemetry (enrollment, responses, etc.) is reported and initial
analyses are conveyed to a clinician. The associated presentations
require very little context - clinicians often have a better
understanding of the data collected than the statistician - often
eliminating the need for narrative components. At the same time, a large
number of hierarchical, heterogenous artifacts (tables and multiple
types of plots) need to be conveyed thereby making the manual creation
of \proglang{R} Markdown documents inconvenient.

@@

The rest of this paper is structured as follow. The next section goes
over basic usage and commentary. This section is meant to convey the
basic approach used by the package and shows how to describe an output
document using \pkg{listdown}, create a document, and change how the
presentation of computational components can be specialized using
\pkg{listdown} decorators. With the user accustomed to the package's
basic usage, section 3 describes the design of the package. Section 4
goes over advanced usage of the package including adding initialization
code to and outputted document as well as how to control chunk-level
options. Section 5 provides a simplified case study of how the package
is currently being used in clinical trial reporting. Section 6 concludes
the paper with a few final remarks on the general types of applications
where \pkg{listdown} has been shown effective.

@@

The \texttt{ld} object, along with the computational components in the
\texttt{comp-comp.rds} file are sufficient to to create the sections,
subsections, and R chunks of a document. The only other thing requires
to create the document is the header. The listdown package currently
supports regular R Markdown and \pkg{workflowr} as \code{yml} objects
from the \pkg{yaml} package \citep{yaml}. These objects are stored as
named lists in R and are easily modified to accomodate document
parameters. A complete document can then be written to the console using
the code shown below. It could easily be written to a file for rendering
using the \texttt{writeLines()} function.

@@

\begin{CodeInput}
R> doc <- c(
R+   as.character(ld_rmarkdown_header("Anscombe's Quartet",
R+                                    author = "Francis Anscombe",
R+                                    date = "1973")),
R+   ld_make_chunks(ld))
R> 
R> doc
\end{CodeInput}

@@

\begin{CodeInput}
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), 
R+                package = "ggplot2",
R+                echo = FALSE)
R> 
R> #cat(paste(ld_make_chunks(ld)[1:7], collapse = "\n"))
R> ld_make_chunks(ld)[1:7]
\end{CodeInput}

@@

While the programmtic generation of reproducible documents has appealing
qualities, it also has fundamental limitations that should be kept in
mind when deciding if tools like \pkg{listdown} should be employed.
First and foremost, without narrative components a document has very
little context. Quantitative analyses require research questions,
hypotheses, reviews, interpretations, and conclusions. Computational
components are necessary but generally not sufficient for constructing
an analysis. This means that if narrative components must be conveyed in
a document, then \pkg{listdown} may make the generation of their
presentation more convenient. However, it does not relieve the burden of
the author to create prose developing a narrative.

@@

\begin{CodeInput}
R> ---
R> title: Simple Trial Report
R> output: html_document
R> ---
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> library(gtsummary)
R> library(flextable)
R> library(DT)
R> library(ggplot2)
R> 
R> cc_list <- readRDS("surv-cc.rds")
R> \```
R> 
R> # Table 1
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> cc_list$`Table 1`
R> \```
R> 
R> # Survival Plots
R> 
R> ## Overall {.tabset}
R> 
R> ### Plot
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, 
R>      fig.height = 6}
R> cc_list$`Survival Plots`$`Overall {.tabset}`$Plot
R> \```
R> 
R> ### Data
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> datatable(cc_list$`Survival Plots`$`Overall {.tabset}`$Data)
R> \```
R> 
R> ### Table
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> datatable(cc_list$`Survival Plots`$`Overall {.tabset}`$Table)
R> \```
R> 
R> ## Stage {.tabset}
R> 
R> ### Plot
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, 
R>      fig.height = 6}
R> cc_list$`Survival Plots`$`Stage {.tabset}`$Plot
R> \```
R> 
R> ### Data
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> datatable(cc_list$`Survival Plots`$`Stage {.tabset}`$Data)
R> \```
R> 
R> ### Table
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> datatable(cc_list$`Survival Plots`$`Stage {.tabset}`$Table)
R> \```
R> 
R> ## Grade {.tabset}
R> 
R> ### Plot
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, 
R>      fig.height = 6}
R> cc_list$`Survival Plots`$`Grade {.tabset}`$Plot
R> \```
R> 
R> ### Data
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> datatable(cc_list$`Survival Plots`$`Grade {.tabset}`$Data)
R> \```
R> 
R> ### Table
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> datatable(cc_list$`Survival Plots`$`Grade {.tabset}`$Table)
R> \```
\end{CodeInput}
\end{CodeChunk}

>>>>

\author{
Michael Kane\\Yale University \And Xun (Tony) Jiang\\Amgen
Inc \And Simon Urbanek\\The University of Aukland
}
\title{On the Programmatic Generation of Reproducible Documents}

\Plainauthor{Michael Kane, Xun (Tony) Jiang, Simon Urbanek}
\Plaintitle{On the Programmatic Generation of Reproducible Documents}
\Shorttitle{On the Programmatic Generation of Reproducible Documents}

\Abstract{
Reproducible document standards, like \proglang{R} Markdown, facilitate
the programmatic creation of documents whose content is itself
programmatically generated. While these documents are generally not
complete in the sense that they will not include prose content,
generated by an author to provide context, a narrative, etc.,
programmatic generation can provide substantial efficiencies for
structuring and constructing documents. This paper explores the
programmatic generation of reproducible by distinguishing components
that can be created by computational means from those requiring
human-generated prose, providing guidelines for the generation of these
documents, and identifying a use case in clinical trial reporting. These
concepts and use case are illustrated through the \pkg{listdown} package
for the \proglang{R} programming environment, which is is currently
available on the Comprehensive R Archive Network (CRAN).
}

@@

      Xun (Tony) Jiang\\
  Amgen Inc\\
  One Amgen Center DriveThousand Oaks, CA 91320-1799, USA\\
  E-mail: \href{mailto:xunj@amgen.com}{\nolinkurl{xunj@amgen.com}}\\

@@

The \pkg{listdown} package provides functions to programmatically create
\proglang{R} Markdown files from named lists. It is intended for data
analysis pipelines where the presentation of the results is separated
from their creation. For this use case, a data processing (or analysis)
is performed and the results are provided in a single named list,
organized hierarchically. List element names denote sections,
subsections, subsubsection, etc. and the list elements contain the data
structure to be presented including graphs and tables. The package has
native support for \pkg{workflowr} \citep{blischak2019}, pdf, word, or
html document along with functions allowing a user to easily extend to
other types of supported documents. The goal of the package is to create
a documents with all tables and visualization that will appear
(computational components). This serves as a starting point from which a
user can organize outputs, describe a study, discuss results, and
provide conclusions (narrative components).

@@

One area where we have found \pkg{listdown} is particularly useful is in
the reporting and research of clinical trial data. These collaborations
tend to be between (bio)statisticians and clinicians either analyzing
past trial data to formulate a new trial or in trial monitoring where
trial telemetry (enrollment, responses, etc.) is reported and initial
analyses are conveyed to a clinician. The associated presentations
require very little context - clinicians often have as good an
understanding of the data collected as that of the statistician's -
often eliminating or significantly reducing the need for narrative
components. At the same time, a large number of hierarchical,
heterogenous artifacts (tables and multiple types of plots) need to be
conveyed thereby making the manual creation of \proglang{R} Markdown
documents inconvenient.

@@

The rest of this paper is structured as follows. The next section goes
over basic usage and commentary. This section is meant to convey the
basic approach used by the package and shows how to describe an output
document using \pkg{listdown}, create a document, and change how the
presentation of computational components can be specialized using
\pkg{listdown} decorators. With the user accustomed to the package's
basic usage, section 3 describes the design of the package. Section 4
goes over advanced usage of the package including adding initialization
code to and outputted document as well as how to control chunk-level
options. Section 5 provides a simplified case study of how the package
is currently being used in clinical trial reporting. Section 6 concludes
the paper with a few final remarks on the general types of applications
where \pkg{listdown} has been shown effective.

@@

The \texttt{ld} object, along with the computational components in the
\texttt{comp-comp.rds} file are sufficient to to create the sections,
subsections, and R chunks of a document. The only other thing requires
to create the document is the header. The listdown package currently
supports regular R Markdown and \pkg{workflowr} as \code{yml} objects
from the \pkg{yaml} package \citep{yaml}. These objects are stored as
named lists in R and are easily modified to accomodate document
parameters. A complete document can then be written to the console using
the code shown below. It could easily be written to a file for rendering
using the \texttt{writeLines()} function, for example.

@@

\begin{CodeInput}
R> doc <- c(
R>   as.character(ld_rmarkdown_header("Anscombe's Quartet",
R>                                    author = "Francis Anscombe",
R>                                    date = "1973")),
R>   ld_make_chunks(ld))
R> 
R> # Write the document.
R> writeLines("anscome-example.rmd")
R> 
R> # Show the document.
R> doc
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), 
R+                package = "ggplot2",
R+                echo = FALSE)
R> 
R> # Show the updated chunk where the library is loaded and data is read.
R> ld_make_chunks(ld)[1:7]
\end{CodeInput}

@@

While the programmtic generation of reproducible documents has appealing
qualities, it also has fundamental limitations that should be kept in
mind when deciding if tools like \pkg{listdown} should be employed.
First and foremost, without narrative components a document has very
little context. Quantitative analyses require research questions,
hypotheses, reviews, interpretations, and conclusions. Computational
components are necessary but generally not sufficient for constructing
an analysis. This means that if narrative components must be conveyed in
a document, then \pkg{listdown} may make the generation of their
presentation more convenient. Narrative components can even by created
in \pkg{listdown} by including \code{character} elements with a chunk
option decorator setting \code{results = "as.is"}. However, it does not
relieve the burden of the author to create prose developing a narrative.

@@

\begin{CodeInput}
R> ---
R> title: Simple Trial Report
R> output: html_document
R> ---
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> library(gtsummary)
R> library(flextable)
R> library(DT)
R> library(ggplot2)
R> 
R> cc_list <- readRDS("surv-cc.rds")
R> \```
R> 
R> # Table 1
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> cc_list$`Table 1`
R> \```
R> 
R> # Survival Plots
R> 
R> ## Overall {.tabset}
R> 
R> ### Plot
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, 
R>      fig.height = 6}
R> cc_list$`Survival Plots`$`Overall {.tabset}`$Plot
R> \```
R> 
R> ### Data
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> datatable(cc_list$`Survival Plots`$`Overall {.tabset}`$Data)
R> \```
R> 
R> ### Table
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5}
R> datatable(cc_list$`Survival Plots`$`Overall {.tabset}`$Table)
R> \```
R> 
R> ## Stage {.tabset}
R> 
R> ### Plot
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, 
R>      fig.height = 6}
R> cc_list$`Survival Plots`$`Stage {.tabset}`$Plot
R> \```
R> 
R> ### Data
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> datatable(cc_list$`Survival Plots`$`Stage {.tabset}`$Data)
R> \```
R> 
R> ### Table
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> datatable(cc_list$`Survival Plots`$`Stage {.tabset}`$Table)
R> \```
R> 
R> ## Grade {.tabset}
R> 
R> ### Plot
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, 
R>      fig.height = 6}
R> cc_list$`Survival Plots`$`Grade {.tabset}`$Plot
R> \```
R> 
R> ### Data
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> datatable(cc_list$`Survival Plots`$`Grade {.tabset}`$Data)
R> \```
R> 
R> ### Table
R> 
R> \```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, 
R>      fig.height = 4.5}
R> datatable(cc_list$`Survival Plots`$`Grade {.tabset}`$Table)
R> \```
\end{CodeInput}
\end{CodeChunk}
